---
title: Practical Skills for<br>Biology Research II
subtitle: Introduction
author: Niklas Mähler
date: 2021-01-01
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("lecture1")
```

```{r packages, include = FALSE, cache = FALSE}
library(maps)
```

# Who am I?

<div class="multicol">
<div class="small" style="width: 40%">

- Niklas Mähler
- Born in Graninge, Sweden in 1986
- Master in Plant Biotechnology at Umeå University 2012
- PhD in Bioinformatics at the Norwegian University of Life Sciences 2016
- Research engineer at UPSC since 2016
- Free time activities
    - Cycling
    - Video games
    - Programming
</div>

<div>
```{r map, fig.width = 6, dev.args=list(bg="transparent")}
nordic_map <- map_data("world", region = c("sweden", "norway", "finland",
                                           "denmark", "germany", "poland",
                                           "estonia", "latvia", "lithuania",
                                           "russia", "belarus"))

locations <- tibble(
  name = c("Graninge", "Umeå", "Ås"),
  lat = c(63.07,  63.80, 59.67),
  long = c(16.94, 20.21, 10.77)
)

ggplot(nordic_map, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group), fill = "steelblue", colour = "black") +
  geom_point(data = locations, fill = "red", colour = "white", size = 5, pch = 21) +
  geom_text(data = locations, aes(label = name), colour = "white",
            hjust = 1, nudge_x = -1, nudge_y = 0.1, size = 6) +
  coord_map(xlim = c(0, 40), ylim = c(55, 71)) +
  labs(x = "Longitude", y = "Latitude") +
  theme_reveal_night()
```
</div>
</div>

# What are we going to look at?

# Data!

- What is data?
- How can we format data?
- How to deal with big datasets?

# Source code!

```{r source_code_example, echo = TRUE, eval = FALSE}
library(tidyverse)
library(maps)

nordic_map <- map_data("world",
                       region = c("sweden", "norway", "finland",
                                  "denmark", "germany", "poland",
                                  "estonia", "latvia", "lithuania",
                                  "russia", "belarus"))

locations <- tibble(
  name = c("Graninge", "Umeå", "Ås"),
  lat = c(63.07, 63.80, 59.67),
  long = c(16.94, 20.21, 10.77)
)

ggplot(nordic_map, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group),
               fill = "steelblue",
               colour = "#CCCCCC") +
  geom_point(data = locations,
             fill = "red", colour = "white",
             size = 5, pch = 21) +
  geom_text(data = locations, aes(label = name),
            colour = "white",
            hjust = 1.3, vjust = 0.3,
            size = 12) +
  coord_map(xlim = c(0, 40), ylim = c(55, 71)) + 
  labs(x = "Longitude", y = "Latitude") +
  theme_reveal_night()
```

# Figures!

- Best practices
- Accessibility

# Sharing stuff!

- How to share code
- How to ask for help
- Transparent research

# What do I expect you to know by the end of this course?

- Being able to use basic R commands to
    - parse data that you have generated
    - transform data to fit whatever downstream analysis you want to perform
    - perform various statistical test (this is not a statistics course though)
- Be able to describe what tidy data is and why this is something we would want to achieve
- Know what version control systems are used for and be able to describe pros and cons with using these

# Recommended literature

<div class="multicol stretch" style="justify-content: space-after;">
<div style="min-width: 60%;">
- R for Data Science<br><https://r4ds.had.co.nz/>
- R graphics cookbook<br><https://r-graphics.org/>
- Advanced R<br><https://adv-r.hadley.nz/>
- ggplot2<br><https://ggplot2-book.org/>
</div>
```{r}
tags$div(
  style = "display: flex; align-items: center; justify-content: space-evenly; flex-wrap: wrap; height: 100%; width: 100%;",
  tags$img(
    src = "https://d33wubrfki0l68.cloudfront.net/b88ef926a004b0fce72b2526b0b5c4413666a4cb/24a30/cover.png",
    style = "max-height: 40%"),
  tags$img(
    src = "https://r-graphics.org/cover.jpg",
    style = "max-height: 40%"),
  tags$img(
    src = "https://d33wubrfki0l68.cloudfront.net/565916198b0be51bf88b36f94b80c7ea67cafe7c/7f70b/cover.png",
    style = "max-height: 40%"),
  tags$img(
    src = "https://ggplot2-book.org/cover.jpg",
    style = "max-height: 40%"),
)
```
</div>

# A brief history of R

# Why R?

R and Rstudio is extremely versatile

All of the presentations that you will see during this course are made in Rstudio.
