---
title: Practical Skills for<br>Biology Research II
subtitle: RStudio
author: Niklas Mähler
date: 2021-01-01
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, cache = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("lecture2")
```

# What is RStudio? { data-transition="slide-in fade-out" }

- Integrated Development Environment (IDE)
- Basically an advanced text editor that knows
  - The languague you are writing in
  - Can run the code that you write, interactively
  - Can display any plots that you generate
  - Can display help for any functions in the language
  - And much, much more

# What is RStudio? { data-transition="fade" }

```{r}
tags$figure(
  tags$img(src = base64_img(here("img/rstudio_screenshot.png"))),
  tags$figcaption("Screenshot of my RStudio window when writing this slide")
)
```


# What is RStudio? { data-transition="fade-in slide-out" }

- Open source
- Free as in free speech, not as in free beer
  - Free version more than enough for our needs though
- RStudio doesn't bundle R itself, this has to be installed separately

# Install R

- Main website: <https://r-project.org>

- Windows:<br><https://ftp.acc.umu.se/mirror/CRAN/bin/windows/base/R-4.0.5-win.exe>
- Mac (High Sierra or later):<br><https://ftp.acc.umu.se/mirror/CRAN/bin/macosx/R-4.0.5.pkg>
- Linux (you know what you're doing):<br><https://ftp.acc.umu.se/mirror/CRAN/bin/linux/>

If you can't find a version for your system, go to the main site and navigate from there.

# Install RStudio

- Main website: <https://www.rstudio.com/>

Go to <https://www.rstudio.com/products/rstudio/download/#download> and download the installer for your platform, run it and follow the instructions.

# Customising RStudio

RStudio is very customisable, both when it comes to appearances and function.
Take some time to set it up to your liking.
This will make the whole experience more enjoyable.