---
title: "Basics of R coding"
subtitle: "Practical Skills for Biology Research II"
author: "Niklas Mähler"
institute: "Umeå University"
date: "2021-10-04"
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("basics_of_r_coding")
```

# Learning goals

- Know about the basic operators in R
- Know what a function is
- Know about the basic data types and structures in R
- Be able to interpret code written by someone else

---

# What is programming?

- Writing instructions telling the computer what to do
- Telling *others* what you want to do.

???

Not only are you telling the computer what you want to do, but you also communicate to others reading your code what you want to do.

---

# R can be a calculator

```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

---

# Common operators

```{r, echo = FALSE}
tribble(
  ~Operator, ~Description,            ~Example,
  "`+`",     "Addition",              "`5 + 2 == 7`",
  "`-`",     "Subtraction",           "`5 - 2 == 3`",
  "`*`",     "Multiplication",        "`5 * 2 == 10`",
  "`/`",     "Division",              "`5 / 2 == 2.5`",
  "`^`",     "Power",                 "`5 ^ 2 == 25`",
  "`%%`",    "Modulo",                "`5 %% 2 == 1`",
  "`%/%`",   "Integer division",      "`5 %/% 2 == 2`"
) %>% knitr::kable()
```

???

Just running these examples in the console, R will print the result.
What if we want to store that result in order to use it somewhere else?

---

# Variable assignment

Variable assignment uses the `<-` operator.

```{r}
x <- 5 + 2
```

This will save the result of the operation into an object `x`.
You can read it as "x gets the value 5 + 2".
This will make it possible to reuse this result later.

```{r}
y <- x ^ 2
```

In order to see what a variable contains, we can simply enter it into the console.

```{r}
y
```


---

# Naming variables

Variables must start with a letter, and they can only contain letters, numbers, and the characters `.` and `_`.

.note.yellow[
`r note_fa("lightbulb")`
If you are familiar with other programming lanugages, you might think that `.` has a special meaning, like in Python where it is used for accessing methods of objects, for example.
In R it's just any old character.
]

There are many conventions when it comes to variable naming:

```
i_use_snake_case
otherPeopleUseCamelCase
some.people.use.periods
And_aFew.People_RENOUNCEconvention
```

I will use `snake_case` in all my examples, and I recommend you do to.
However, the most important thing is **consistency**.

.note.red[
`r note_fa("exclamation-triangle")`
Note that R is picky.
Spelling and character case matters!
There is a difference between `myVariable` and `myvariable`.
]

---

# Use meaningful variable names

What is easier to understand? This:

```{r}
a <- 100
b <- 50
c <- 5
d <- rnorm(a, b, c)
```

or this:

```{r}
number_of_samples <- 100
sample_mean <- 50
sample_standard_deviation <- 5
random_samples <- rnorm(number_of_samples, sample_mean, sample_standard_deviation)
```

???

These two blocks of code do the exact same thing, but one of them also tries to **explain** what it is doing by having meaningful variable names.
Sure, it's much more typing, but you will be happy when you revisit your code after a while and can pick up where you left off.

---

# Variable comparisons

Often in programming we want to compare values, e.g. "is x larger than y?", or "is x equal to y?".

```{r, echo = FALSE}
tribble(
  ~Operator, ~Description,            ~Example,
  "`<`",     "Strictly less than",    "`5 < 7`",
  "`>`",     "Strictly greater than", "`7 > 5`",
  "`<=`",    "Less or equal to",      "`7 <= 7`",
  "`>=`",    "Greater or equal to",   "`7 >= 6`",
  "==",      "Equals",                "`7 == 7`",
  "!=",      "Not equals",            "`7 != 5`"
) %>% knitr::kable()
```

All of the examples in the table evaluate to `TRUE`.

---

# Comparisons of decimal numbers

.note.red[
`r note_fa("exclamation-triangle")`
When doing exact comparisons of decimal numbers, you might get unexpected results.
This is due to the inability of computers to represent decimal numbers accurately, something called [floating point precision](https://en.wikipedia.org/wiki/Floating-point_arithmetic).
]

```{r}
(0.3 * 3) + 0.1 - 1
```

--

```{r}
(0.3 * 3) + 0.1 == 1
```

--

```{r}
all.equal((0.3 * 3) + 0.1, 1)
```

???

So, we see here that we have different representations of variables in R.
Decimal numbers, or floating point numbers, are different from integer numbers, and we have also seen that we have a `TRUE`/`FALSE`.

---

# Calling functions

.left-column[
## What is a function?
]

.right-column[
> In computer programming, a subroutine is a sequence of program instructions that performs a specific task, packaged as a unit.
> This unit can then be used in programs wherever that particular task should be performed.

.right[https://en.wikipedia.org/wiki/Subroutine]

You have already used quite a few functions, for example when visualising data with ggplot2.
R has a lot of built-in functions that we tend to rely more or less upon.
]

???

This is a general concept, and nothing that is unique for R.

---

# Calling functions

.left-column[
## What is a function?
## Structure
]

.right-column[
Functions are structured like this:

```{r, eval = FALSE}
function_name(arg1 = val1, arg2 = val2, ...)
```

.note.yellow[
`r note_fa("lightbulb")`
When typing the name of a function, you can press the <kbd>Tab</kbd> key on your keyboard in order to get autocompletion suggestions.
This also works for variables, and it's a function I use all the time.
Try to make it a habit, since it also helps you avoid simple mistakes, such as spelling errors.
]
]

---

# Calling functions

.left-column[
## What is a function?
## Structure
## Function arguments
]

.right-column[
Arguments can be supplied either by name, or by position, or a combination of the two.
For example, the function `seq` has the following definition:

```{r, eval = FALSE}
seq(from = 1, to = 1, by = ((to - from)/(length.out - 1)),
    length.out = NULL, along.with = NULL, ...)
```
]

???

Let's use the `seq` function as an example.
This function create sequences of numbers based on the input arguments.

Whenever an argument is followed by an equals sign and an expression, this is the default value for this argument.
Here we can see that the default value of `from` is 1, and the default value of `to` is also 1.
What do you think the result would be if we called this function without any arguments at all?

--

.right-column[
And we can call it in a number of different ways:

```{r}
seq(0, 10, 2)
```
]

--

.right-column[
```{r}
seq(from = 0, to = 10, by = 2)
```
]

--

.right-column[
```{r}
seq(0, 10, length.out = 6)
```
]

???

All of these calls produce the same result, but in slightly different ways.
The important thing to take away from this is that if we give a function unnamed arguments, the order that we give them must match with the order in the function definition.

Later in the course we will go over how we can write our own functions.

---

# Atomic data types

.left-column[
## Double
]

.right-column[
Double-precision floating point values, "regular numbers".

```{r}
pi
typeof(pi)
typeof(5)
```
]

???

As the name implies, atomic data types refer to the smallest types that we have in R.
From these we can build more complex structures.

---

# Atomic data types

.left-column[
## Double
## Integer
]

.right-column[
Numbers without any decimal fraction.
These we normally don't come across in everyday analyses, at least not in their pure form, since numbers normally are represented as doubles.

```{r}
5L
typeof(5L)
```
]

---

# Atomic data types

.left-column[
## Double
## Integer
## Character
]

.right-column[
Text, or strings.
Has to be enclosed in quotes.

```{r, eval = FALSE}
x <- "hello world"
```

If we need to represent quotes in a character string, we can escape them with a backslash:

```{r}
x <- "hello \"world\""
```
]

???

Both single and double quotes work, but preferably don't mix them.

If we don't use quotes, this is something called a "bare name", and R will in most cases try to find an object with that name in your environment.
I will not explain the exceptions to this here and now though, just assume that if you see a word without quotes, it is either referring to a reserved keyword in R, or an object.

---

# Atomic data types

.left-column[
## Double
## Integer
## Character
## Logical
]

.right-column[
Boolean values, either `TRUE` or `FALSE`.

```{r}
5 < 7
"R" == "Python"
```
]

---


# Vectors

Everything we've done so far has been using vectors, and this is the default data structure in R.
Vectors are composed of elements of a single data type, and they have a length.

```{r}
x <- 5
length(x)
```

--

We can use the `c` function to **c**ombine objects into a vector.

```{r}
x <- c(1, 2, 3, 4, 5)
length(x)
```

--

There is also a short-hand way of creating numerical vectors as the one above.

```{r}
x <- 1:5
x
```


---

# Vectors

If we try to combine objects of different types, R tries to convert them to a representation common to all of them.

```{r}
c(1, 2, 3L, 4L)
```

--

```{r}
c("one", 2, 3L, "four")
```

--

```{r}
c(0, 1, TRUE, FALSE)
```

--

```{r}
c(0, 1, TRUE, FALSE, "hello")
```

---

# Vectors

The elements in a vector can be named.

```{r}
x <- c(first = 0, second = 1, third = 2)
x
```

--

```{r}
names(x)
```

--

```{r}
names(x) <- c("one", "two", "three")
x
```

---

# Vectors

We can access individual elements in a vector using square brackets.

```{r}
x <- 10:15
x
x[1]
x[2]
```

.note.red[
`r note_fa("exclamation-triangle")`
If you have some experience in programming, you might be used to numbering starting at 0.
This is not the case in R where indexing is 1-based.
]

---

# Vectors

You can also get a range of numbers by indexing a vector with another vector.

```{r}
x <- 10:15
x
x[c(1, 3, 5)]
```

--

Yet another way of indexing is by using a boolean vector.

```{r}
x[c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)]
```

---

# Vectors

Finally, if your vector has names, you can also access elements by using these.

```{r}
x <- c(first = 0, second = 1, third = 2, fourth = 3)
x
x["first"]
```

--

Just as when indexing with numbers, you can use a vector of strings.

```{r}
x[c("first", "third")]
```

---

# Missing values

Sometimes you want to be able to represent missing values.
It could be that a particular experiment failed, or that a measurement was outside of the detection range of your instrument.

These values can be represented with `NA`.

```{r}
x <- c(1, 2, 3, NA, 5)
```

???

It is an important distinction between a missing value, and just setting a value to a particular value, e.g. zero.
Was the value of the measurement really zero, or is that data point simply missing?

---

# Missing values

```{r}
x <- c(1, 2, 3, NA, 5)
mean(x)
mean(x, na.rm = TRUE)
(x[is.na(x)] <- 0)
mean(x)
```

???

Many functions take missing values into consideration, and then it is also important to think about how these should be handled.
Oftentimes, if you try to calculate something on a vector containing missing values, you will get a missing value as the result.
You then have the option of excluding those values.
Note that this is not equivalent of setting the missing value to a fixed value.

---

# Composite data types

.left-column[
## Factor
]

.right-column[
Used for representing categorical variables.

```{r}
factor(c(1, 1, 2, 2, 3, 3))
```
]

--

.right-column[
Each level can also be associated with a label.
By default, the label is the same as the original value.

```{r}
factor(c(1, 1, 2, 2, 3, 3), labels = c("one", "two", "three"))
```
]

--

.right-column[
... but the underlying representation is still integers, regardless of the original data type.

```{r}
as.numeric(factor(c("one", "one", "two", "two", "three", "three")))
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
]

.right-column[
A matrix can be seen as a two-dimensional vector.
Similar to vectors, the matrix must contain a single data type.

```{r}
matrix(1:9, nrow = 3)
```

Just like vectors, it can have names, but instead of every single element being associated with a name, each row and column get a name.

```{r}
x <- matrix(1:9, nrow = 3)
rownames(x) <- c("row1", "row2", "row3")
colnames(x) <- c("col1", "col2", "col3")
x
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
]

.right-column[
Matrix indexing is similar to vector indexing, but we now have two dimensions to consider.
We then refer separately to rows and columns.

```{r}
x <- matrix(1:9, nrow = 3)
x
row <- 1
col <- 2
x[row, col]
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
]

.right-column[
If one of them is left blank, it will return all elements in that row or column.

```{r}
x
x[2, ]
x[, 2]
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
]

.right-column[
Contrary to vectors and matrices, lists can contain objects of different types.
These elements can also be named.

```{r}
x <- list(one = 1, two_vector = 1:3,
          three_matrix = matrix(1:9, nrow = 3))
x
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
]

.right-column[
Printing a very big list can become quite messy.
In order to get an overview of what the list contains, without printing every single element and subelement, we can look at the **str**ucture of the list with the `str` function.

```{r}
str(x)
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
]

.right-column[
Indexing a list is a bit different from indexing a vector.
If we use single square brackets, we get a list back.

```{r}
x <- list(one = 1:3, two = letters[1:5], three = rnorm(4))
x[1]
typeof(x[1])
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
]

.right-column[
If we want the underlying object only, then use double brackets.

```{r}
x[[1]]
typeof(x[[1]])
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
]

.right-column[
If the elements are named, we can index by name just as well as position.

```{r}
x["one"]
x[["one"]]
```

... and we can also use the `$` operator.

```{r}
x$one
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
]

.right-column[
Can be seen as a special case of a list where the length of each element is the same.

```{r}
x <- data.frame(some_letters = letters[1:5], x = 6:10, y = rnorm(5))
x
```
]

???

This is probably the data structure that we will be spending most of our time with, at least during this course.

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
]

.right-column[
Indexing a data frame has elements of all the different indexing methods we've seen so far.
We can use square brackets to access rows and columns.

```{r}
# Element in first row, second column
x[1, 2]
# First row
x[1, ]
# Second column
x[, 2]
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
]

.right-column[
We can also use the `$` operator to access individual columns.

```{r}
x$some_letters
```
]

--

.right-column[
... which in turn is a vector that we can index.

```{r}
x$some_letters[2]
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
## Tibble
]

.right-column[
For all practical purposes, this is a data frame, but with some extra bells and whistles.
The datasets that we looked at last week (`mpg` and `diamonds`) were both tibbles.
Anything you can do on a data frame, you can do on a tibble.

.note.red[
`r note_fa("exclamation-triangle")`
Some packages do overly strict checks on data frames, and might cause some things to fail if you supply a tibble instead of a traditional data frame.
Fortunately, this is easy to get around.
Just convert your tibble to a data frame: `as.data.frame(your_tibble)`
]
]

???

I would argue that there are very few reasons to using traditional data frames instead of tibbles.
There are some cases where you are kind of forced to use data frames, but these situations are few and far between.
All of the tools that we will be presenting during these weeks work perfectly well with tibbles.

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
## Tibble
]

.right-column.small[
```{r}
mpg
```
]

---

# Composite data types

.left-column[
## Factor
## Matrix
## List
## Data frame
## Tibble
]

.right-column.small[
```{r}
as.data.frame(mpg)
```
]

???

Not only does tibbles print better, but they also have functionality that can help us easier make sense of our data.
We will go into these things later this week.