---
title: Practical Skills for<br>Biology Research II
subtitle: Installing RStudio
author: Niklas Mähler
date: 2021-01-01
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, cache = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("lecture3")
```

# Learning goals

- Know where to find documentation for functions
- Can describe what an R package is

# Getting help

```{r, echo = TRUE, attr.source=".numberLines"}
?mean
```

```{r}
tags$img(src = base64_img(here("img/help_example_screenshot.png")),
         class = "stretch")
```

# Anatomy of a help page { data-transition="slide-in none" }

```{r}
tags$img(src = base64_img(here("img/help_example_screenshot.png")),
         class = "stretch")
```

# Anatomy of a help page { data-transition="none" }

```{r}
tags$img(src = base64_img(here("img/help_example_screenshot.png")),
         class = "stretch")
```

# Anatomy of a help page

<div class="smaller left-justify stretch" style="background-color: #333; padding: 1em; margin: 0 auto; overflow: scroll;">

<p style="margin-top: 0;">mean {base}</p>

## Arithmetic Mean

### Description

Generic function for the (trimmed) arithmetic mean.

### Usage

<pre style="width: 100%; margin-top: 0; font-size: 1em;"><code>mean(x, ...)

## Default S3 method:
mean(x, trim = 0, na.rm = FALSE, ...)</code></pre>

### Arguments

```{r}
tags$div(
  tags$div(
    class = "row",
    style = "display: flex; margin-top: 0.5em;",
    tags$div(
      tags$code("x"),
      style = "min-width: 30%;"
    ),
    tags$div(
      tags$p(
        "An", tags$code("R"), "Object. Currently there are methods for numeric/logical vectors and date, date-time and time interval objects. Complex vectors are allowed for ", tags$code("trim = 0"), ", only.",
        style = "margin: 0;")
    )
  ),
  tags$div(
    class = "row",
    style = "display: flex; margin-top: 0.5em;",
    tags$div(
      tags$code("trim"),
      style = "min-width: 30%;"
    ),
    tags$div(
      tags$p(
        "the fraction (0 to 0.5) of observations to be trimmed from each end of ", tags$code("x"), " before the mean is computed. Values of trim outside that range are taken as the nearest endpoint.",
        style = "margin: 0;")
    )
  ),
  tags$div(
    class = "row",
    style = "display: flex; margin-top: 0.5em;",
    tags$div(
      tags$code("na.rm"),
      style = "min-width: 30%;"
    ),
    tags$div(
      tags$p(
        "a logical value indicating whether ", tags$code("NA"), " values should be stripped before the computation proceeds.",
        style = "margin: 0;")
    )
  ),
  tags$div(
    class = "row",
    style = "display: flex; margin-top: 0.5em;",
    tags$div(
      tags$code("..."),
      style = "min-width: 30%;"
    ),
    tags$div(
      tags$p(
        "further arguments passed to or from other methods",
        style = "margin: 0;")
    )
  )
)
```

### Value

If `trim` is zero (the default), the arithmetic mean of the values in `x` is computed, as a numeric or complex vector of length one. If `x` is not logical (coerced to a numeric), numeric (including integer) or complex, `NA_real_` is returned, with a warning.

If `trim` is non-zero, a symmetrical trimmed mean is computed with a fraction of `trim` observations deleted from each end before the mean is computed.

### References

Becker, R. A., Chambers, J. M. and Wilks, A. R. (1988) *The New S Language*. Wadsworth & Brooks/Cole.

### See Also

`weighted.mean`, `mean.POSIXct`, `colMeans` for row and column means.

### Examples

<pre style="width: 100%; margin-top: 0; font-size: 1em;">x <- c(0:10, 50)
xm <- mean(x)
c(xm, mean(x, trim = 0.10))</code></pre>

</div>