---
title: "Data sharing"
subtitle: "Practical Skills for Biology Research II"
author: "Niklas Mähler"
institute: "Umeå University"
date: "2021-10-14"
editor_options:
  chunk_output_type: console
---

layout: true

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("data_sharing", filename = "17_data_sharing.Rmd")
library(readxl)
```

---

# Why do we need to share our data?

- Science is a community effort
- We need to be transparent about our research

---

```{r, echo = FALSE}
articles <- list(
  article_tile(
    "Scientists Have a Sharing Problem",
    "Competition and disorganization within their disciplines prevent many researchers from making their data publicly available, which is stunting scientific progress.",
    "2014-12-15",
    "The Atlantic",
    "https://www.theatlantic.com/health/archive/2014/12/scientists-have-a-sharing-problem/383061/"
  ),
  article_tile(
    "“Available upon request”: not good enough for microbiome data!",
    "",
    "2018-01-10",
    "Microbiome",
    "https://doi.org/10.1186/s40168-017-0394-z"
  ),
  article_tile(
    "Assessing data availability and research reproducibility in hydrology and water resources",
    "Bottlenecks [for reproducibility] include: only some digital artifacts available (44% of articles), no directions (89%), or all artifacts available but results not reproducible (5%).",
    "2019-02-26",
    "Scientific Data",
    "https://doi.org/10.1038/sdata.2019.30"
  ),
  article_tile(
    "High profile coronavirus retractions raise concerns about data oversight",
    "Retracted studies had relied on health-record analyses from a company that declined to share its raw data for an audit.",
    "2020-06-05",
    "Nature",
    "https://doi.org/10.1038/d41586-020-01695-w"
  ),
  article_tile(
    "Why the United States is having a coronavirus data crisis",
    "Political meddling, disorganization and years of neglect of public-health data management mean the country is flying blind.",
    "2020-08-25",
    "Nature",
    "https://doi.org/10.1038/d41586-020-02478-z"
  ),
  article_tile(
    "Data sharing practices and data availability upon request differ across scientific disciplines",
    "",
    "2021-07-27",
    "Scientific Data",
    "https://doi.org/10.1038/s41597-021-00981-0"
  ),
  class = "article-wrapper"
)

set.seed(5)
collect_articles(articles, rotation = 5)
```

???

If we think about reproducibility, there are countless examples of studies where analyses cannot be reproduced due to data not being available.

The COVID-19 pandemic has put some of these issues in the spotlight, where data being held back, or getting lost, has resulted in retracted studies and 
Not only is this problematic from a reproducibility point of view, but these are matters of public health.

The first study in Scientific Data shows that out of the papers that they looked at 

The second study published in Scientific Data looked at over 800 papers published in Science and Nature in to time periods: 2000-2009 and 2010-2019.
They could see that data availability has improved with time, but there is still work to do.
In cases where data was missing, or partially missing, they contacted the authors, and 41% of these requests were ignored within 60 days, including two reminders.

---

# A community effort

- Allowing others to scrutinise and expand on your work
- Reviews
  - It might often be relevant to reanalyse part of the data from the studies included.
- Meta-studies
  - Examples in humans are [identifying longevity genes](https://doi.org/10.1038/s41467-019-11558-2), [identifying genetic variation associated with human height and BMI](https://doi.org/10.1093/hmg/ddy271), and [associating genetic variation with the gut microbiota](https://doi.org/10.1038/s41588-020-00763-1).

---

# Concerns about data sharing

- Not knowing where to put the data.

???

This is something I hope you will know more about after this lecture.

--
- Lack of time.

???

This is not something I can help you with right now, but the solution is quite easy: take this into account when planning a project.
Nowadays, this is an essential part of science, and it should be in your time plan and your budget just as much as anything else.

--
- Concerns about privacy and confidentiality

???

Getting scooped is perhaps the most common concern when it comes to sharing data.
There are also ways around this.
For example, many repositories have a function where you can keep the data private until your study has been accepted.
However, the fear of getting scooped is often exaggerated, especially if you are "just" sharing the data.
In order to get scooped, someone else would need to have the same ideas as you, and the same skillset within their team, and they also constantly have to browse data repositories in order to find the data that they need for this, supposedly, specific idea.

--
- You're not the one making the final decision.

???

Often, you are not the one taking the final decision when it comes to publication strategy.
If people in your team oppose themselves to sharing the data, at least voice your opinion if you think it would be a good idea to do so.
Also, as we'll get to know, sharing data doesn't necessarily mean that we have to put it out there for all to see, and this ties back to the previous point.
We can actually be transparent without giving absolutely everything away, and sometimes this is crucial.
For example, if you think about studies involving humans, you have privacy concerns you have to take into account.

---

# The FAIR principles

FAIR was established in 2016, and it is an abbreviation that outlines a set of principles for how digital assets should be handled.

[The FAIR Guiding Principles for scientific data management and stewardship](https://www.nature.com/articles/sdata201618)

---

# The FAIR principles

.left-column[
## Findable
]

.right-column[
First of all, you must be able to find the data that you are looking for.
For this to work well, a number of things are needed:

- Metadata has to exist and be exhaustive enough.
- The metadata has to be available for people to find.
]

---

# The FAIR principles

.left-column[
## Findable
## Accessible
]

.right-column[
Second, you should be able to find information on how to access the data.
This doesn't necessarily imply that the data has to be *publicly* accessible, but rather that any associated metadata should contain information on how to get access to the data.
Any conditions for accessing and reusing the data should be stated as well.
]

---

# The FAIR principles

.left-column[
## Findable
## Accessible
## Interoperable
]

.right-column[
Interoperability is a complex topic which includes

- Use of standards when recording data
  - Dates
  - Abbreviations
  - Ontologies
- Use of file formats
  - They work on as many systems and software as possible
    - Software in turn should also meet some requirements for longevity 
  - They are well documented
  - Should be accessible in the future
  - Are open/non-proprietary
]

---

# The FAIR principles

.left-column[
## Findable
## Accessible
## Interoperable
## Reusable
]

.right-column[
Finally, the data has to be reusable.
This means that the data needs to have metadata that describes the data in detail, including the aim of the original study, and how the data was gathered and processed.
Any licenses that apply to the data must be listed.
]

---

# How FAIR are we?

```{r fair_how_familiar, echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 12}
download.file("https://ndownloader.figshare.com/files/25563296", sod_file <- tempfile(fileext = ".xlsx"))
sod_data <- read_excel(sod_file, skip = 1, sheet = 1)

sod_data %>% 
  select(Year, matches("Thinking about")) %>% 
  rename(year = Year,
         how_familiar = `Thinking about the FAIR principles: Findable, Accessible, Interoperable and Reusable. How familia...`) %>% 
  filter(!is.na(year), !is.na(how_familiar), how_familiar != "0") %>% 
  mutate(how_familiar = str_replace_all(how_familiar, c("I've" = "I have", "I'm" = "I am")) %>% 
           str_wrap(width = 25),
         across(c(year, how_familiar), as.factor),
         how_familiar = fct_inorder(how_familiar),
         year = fct_rev(year)) %>%
  group_by(how_familiar) %>% 
  count(year) %>% 
  group_by(year) %>% 
  mutate(fraction = n / sum(n)) %>% 

  ggplot(aes(how_familiar, fraction, fill = year, group = year)) +
  geom_col(position = position_dodge2()) +
  coord_flip() +
  scale_fill_brewer(palette = 7, type = "qual", guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 0.6),
                     expand = expansion(add = c(0.01, 0.02)), breaks = seq(0, 1, by = 0.2)) +
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_blank(),
        axis.text = element_text(size = 20),
        axis.ticks = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.grid = element_blank(),
        panel.grid.major.x = element_line(colour = "gray"))
```

[State of Open Data survey 2020 data](https://figshare.com/articles/dataset/_/13274744)<br>
[State of Open Data survey 2020 report](https://digitalscience.figshare.com/articles/report/The_State_of_Open_Data_2020/13227875)

???

This is from a survey from the State of Open Data --- a research project whose goal has been to the review state of the open data movement, and how it has progressed for the past 10 years.

---

# How FAIR are we?

One problem with the FAIR principles is that they are quite open for interpretation, which results in different researchers and institutes implementing them differently.

- [Swedish National Data Service](https://snd.gu.se/en/describe-and-share-data/what-does-fair-data-mean)
- [The GO FAIR initiative](https://www.go-fair.org/fair-principles/)
- [State of Open Data](https://stateofopendata.od4d.net)

---

class: title

# Where can we share data?

???

Here I will mostly focus on the types of data that I've been working with.
If you have experience from other types of data, please let me know, and we can see if we can find a good place for making that type of data public.

---

# Sequence data

- [European Nucleotide Archive (ENA)](https://www.ebi.ac.uk/ena/browser/home)
  - Hosted by EMBL-EBI
  - Nucleotide sequence information
  - Raw sequencing data
  - Genome assemblies
  - Functional annotations
- [NCBI](https://www.ncbi.nlm.nih.gov/)
  - Hosted by the National Institutes of Health (NIH)
  - Raw sequencing data
  - Genome assemblies
  - Functional annotations
  
???

NCBI — National Center of Biotechnology Information, hosted by the NIH — National Institutes of Health

This can be many types of sequence data, including things like targeted sequencing technologies like ChIP-Seq and Hi-C.

Both of these are very strict when it comes to what metadata they require.
There is a lot of information you have to collect about your data before you can submit.

---

# Miscellaneous

- [Zenodo](https://zenodo.org)
- [Figshare](https://figshare.com)
- [Dryad](https://datadryad.org)
- [Mendeley Data](https://data.mendeley.com/)


- Basically any type of dataset
- Posters and presentations
- Preprints
- Software

???

It can be raw images for image analysis, sequencing data, random excel sheets, you name it.

Super simple to use.
Puts more responsibility on the user in terms of metadata since you can upload more or less what you want.

---

# Licencing

## What is a licence?

It is a legal tool that defines how someone can modify, share, or otherwise use a resource for which you have the copyright, while keeping the copyright.

## Things to consider

- Attribution
  - Do you have to credit the original creator?
- Copyleft
  - Can the data be modified and reshared, with the same licence?
- Non-commerciality
  - Can you use the resource to make money?


- [choosealicense.com](https://choosealicense.com/) can help you choose a licence, intended mainly for software.
- [Creative Commons](https://creativecommons.org/licenses/) have a set of licences that are appropriate for papers, data, posters, etc.

???

Often when sharing things, be it a poster for a conference, a piece of software we wrote, or a dataset, we have to choose what licence that should apply.
The licence tells anyone that comes across it what the conditions are for using, reusing, or modifying the resource.

If you don't choose a licence, the resource is automatically set to "all rights reserved" according to copyright law, meaning no-one can do anything with it.
That kind of sucks.

---

# Some things to check out

- [Webinar on open access data sharing and submission](https://www.youtube.com/watch?v=VF12vht2qmE)
- [GO FAIR](https://www.go-fair.org/)
- [Creative Commons license chooser](https://creativecommons.org/choose/)