---
title: "Control flow"
subtitle: "Practical Skills for Biology Research II"
author: "Niklas Mähler"
institute: "Umeå University"
date: "2021-10-22"
editor_options:
  chunk_output_type: console
---

layout: true

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
source(here::here("scripts/setup.R"))
setup_presentation("control_flow", filename = "22_control_flow.Rmd")
```

---

# What we'll learn about

- `if` statements
- `switch` statements
- `while` loops
- `for` loops

---

# Choices and loops

There are two ways we can control what parts of our code gets executed: choices and loops.

Choices can be handled by `if` statements, calls to `switch()` or the `ifelse()` function. Based on the value of a variable, we can decide how to proceed in our code.

Loops in R are `for` loops and `while` loops. They are useful whenever we have to repeat the same operation many times.

---

class: title

# Choices

---

# `if`

The `if` statement looks like this in R

```{r, eval = FALSE}
if (condition) true_action else false_action
if (condition1) true_action1 else if (condition2) true_action2 else false_action
```

and we can read it like "if `condition` is true, then evaluate `true_action`, otherwise evaluate `false_action`.

The evaluation branches of the `if` statement can also be written on multiple lines with the help of `{`.

```{r, eval = FALSE}
if (condition) {
  true_action
} else {
  false_action
}
```

---

# Returns from `if`

The if statement returns the last statement of whatever branch was evaluated, so we can use it for variable assignment.

```{r}
x <- 0
state <- if (x < 0) "negative" else "positive"
state
```

This is convenient in functions.

```{r}
grade <- function(score) {
  if (score < 50) {
    "U"
  } else if (score < 80) {
    "G"
  } else {
    "VG"
  }
}
```

???

When we have multiple conditions like this, we have to think about what order we have them in.
The first branch whose condition evaluates to `TRUE` will be executed.

---

# No `else` matters

If we have an `if` statement with a single condition and no `else` statement, `NULL` will be returned.

```{r}
(if (FALSE) 5)
```

This can be helpful in certain functions where `NULL` values are ignored.

```{r}
greet <- function(name, birthday = FALSE) {
  str_c("Hello ", name,
        if (birthday) ", and happy birthday!")
}
greet("Per")
greet("Lisa", TRUE)
```

---

# Invalid inputs

`if` statements only takes a single logical value as input.

```{r, error = TRUE}
if ("x") 5
if (logical(0)) 3
if (NA) 4
```

---

# Invalid inputs

Logical vectors longer than 1 are accepted, but we get a warning.

```{r, warning = TRUE}
if (c(TRUE, FALSE)) 1
```

---

# `if` for vectors

If we have a vector of conditions we can use `ifelse()` instead of `if`.
It is specified as

```{r, eval = FALSE}
ifelse(condition, yes, no)
```

where the arguments will be recycled to the same length.

```{r}
x <- 1:10
ifelse(x %% 5, letters, as.character(x))
```

```{r}
x <- 1:10
ifelse (x %% 2 == 0, "even", "odd")
```

`r note("The <code>yes</code> and <code>no</code> vectors should be of the same type in order to avoid confusion", "warning")`

---

# `switch()`

.pull-left[
Instead of writing code like this
```{r, error = TRUE}
recode_animal <- function(x) {
  if (x == "cow") {
    1
  } else if (x == "chicken") {
    2
  } else if (x == "sheep") {
    3
  } else if (x == "human") {
    4
  } else {
    stop("invalid `x` value")
  } 
}
recode_animal("human")
recode_animal("shark")
```
]

.pull-right[
We can write it like this
```{r, error = TRUE}
recode_animal <- function(x) {
  switch(x,
         cow = 1,
         chicken = 2,
         sheep = 3,
         human = 4,
         stop("invalid `x` value")
  )
}
recode_animal("human")
recode_animal("shark")
```
]

???

It is recommended to always have the last case being an error.
Otherwise it will return `NULL` silently, and this can cause headaches.

Another thing is that while this would work for a numerical value of `x`, it is limited to whole numbers, and it might fail unexpectedly if they are not.
Therefore it is recommended to only use this for character input.

---

class: title

# Loops

---

# `for` loops

A `for` loop can be used to iterate over the values of a vector.
The basic form is

```{r, eval = FALSE}
for (item in vector) perform_action
```

and we can read it like "evaluate `perform_action` for each item in `vector`".
`item` is in this case updated for each turn in the loop to reflect the current value in `vector`.
Just like with `if` statements, we can have `perform_action` on multiple lines with `{`.

```{r}
for (i in 1:3) {
  print(i)
}
```

???

When iterating over a vector of indices, it is common practice to have one-letter names, such as `i`, `j`, or `k`.

---

# Loop environment

It's worth noting that the assignment of `item` in the loop happens in the current environment, so it will overwrite the variable if it already exists.

```{r}
item <- "precious"
for (item in c("fish", "taters", "nasty hobbitses")) {}
item
```

The same is true for the loop body.

---

# Early termination

There are two ways to terminate a `for` loop early: `next` and `break`.

`next` will interrupt the current iteration and immediately continue with the next iteration.

```{r}
for (i in 1:5) {
  if (i <= 3) {
    next
  }
  print(i)
}
```

---

# Early termination

`break` will exit the loop entirely.

```{r}
for (i in 1:1000000) {
  if (i > 5) {
    break
  }
  print(i)
}
```

---

# Pitfalls

.left-column[
## Preallocation
]

.right-column[
If we are using the loop to generate a vector of results, it is recommended to define the vector beforehand, otherwise it will be very slow.

```{r}
x <- list(c(1:10), rnorm(100), runif(100))
means <- double(length(x))
for (i in 1:length(x)) {
  means[i] <- mean(x[[i]])
}
means
```
]

---

# Pitfalls

.left-column[
## Preallocation
## `1:length(x)`
]

.right-column[
Beware of looping over `1:length(x)`.
If `x` has a length of 0, this will produce unexpected results.

```{r}
x <- c()
means <- double(length(x))
for (i in 1:length(x)) {
  means[i] <- mean(x[i])
}
1:length(x)
```

In this case it's better to use `seq_along`.

```{r}
means <- double(length(x))
for (i in seq_along(x)) {
  means[i] <- mean(x[i])
}
means
```
]

???

With `seq_along` we see that the loop isn't run at all.

---

# `while` and `repeat`

`while` loops takes a condition and keeps running the loop while the condition is `TRUE`.

```{r, eval = FALSE}
while (condition) action
```

`repeat` works similarly, but it doesn't take a condition.
It simply loops until `break` is called.

```{r, eval = FALSE}
repeat(action)
```

Any `for` loop can be written as a `while` loop, and any `while` loop can be written as a `repeat`.

There are however other solutions that we can use instead of looping that makes our code more readable, and we'll learn about these methods when talking about vectorisation and functional programming.

???

So, `while` is more flexible than `for` and `repeat` is more flexible than `while`.
I would argue that you should always use a `for` loop whenever you can express an iteration as such, since it's more readable, and fewer cases where things can go wrong.

---

class: title

# Let's improve our functions!

---

# Suggested reading


- R for Data Science
  - [19.4 Conditional Execution](https://r4ds.had.co.nz/functions.html?q=if%20statement#conditional-execution)
  - [21 Iteration](https://r4ds.had.co.nz/iteration.html)
- [Advanced R: Control Flow](https://adv-r.hadley.nz/control-flow.html)